Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2025-04-04 paper/paper.tex
fix clipping in appendix

Next, a parse in the equalizer $\equalizer{a}{f}{g}$ is defined as
a parse in $\sem{A}$ that is mapped to the same parse by the parse transformers $\sem{f}$ and
$\sem{g}$.
%
The universe $L$ of linear types is interpreted as the set of all
small grammars.
\begin{figure}
  \begin{minipage}[t]{.3\textwidth}
  \begin{footnotesize}
    \begin{flalign*}
    &\sem{c}\gamma\,w = \{ c | w = c\}\\
    &\sem{I}\gamma\,w = \{ () \pipe w = \epsilon \}\\
    &\sem{\uparrow A}\gamma = \sem{A}\gamma\,\epsilon\\
    &\sem{A \lto B}\gamma\,w = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,ww'\\
    &\sem{B \tol A}\gamma\,w = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'w\\
    &\sem{\el(F)}\gamma\,G = \sem{F}\gamma\,G\\
    &\sem{\map(F)}\gamma\,f = \sem{F}\gamma\,f\\
    &\sem{\mu A}\gamma = \mu (\sem{A}\gamma)
  \end{flalign*}
  \end{footnotesize}
  \end{minipage}%
  \begin{minipage}[t]{.7\textwidth}
  \begin{footnotesize}
    \begin{flalign*}
    &\sem{A \otimes B}\gamma\,w = \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in \sem{A}\gamma\,w_1 \wedge b \in \sem{B}\gamma\,w_2 \}\\
    &\sem{\bigoplus_{x:X} A}\gamma\,w = \{ (x, a) \pipe x \in \sem{X}\gamma
      \wedge a \in \sem{A}(\gamma,x)w \} \\
    &\sem{\bigwith_{x:X} A}\gamma\,w = \prod_{x\in \sem{X}\gamma} \sem{A}(\gamma, x)\,w \\
    &\sem{\equalizer{a}{f}{g}}\gamma\,w = \{ a \in \sem{A}\gamma\,w \pipe \sem{f}\gamma\,w\,a = \sem{g}\gamma\,w\,a \} \\
    &\sem{L}\gamma = \Grammar_0\\
    &\sem{SPF\,X}\gamma = \textrm{DepPolyFunctor}(\sem{X}\gamma\times \Sigma^*,\Sigma^*)
  \end{flalign*}
  \end{footnotesize}
  \end{minipage}
  \caption{Grammar Semantics}
  \label{fig:semantics}
\end{figure}
The most complex part of the semantics is the interpretation of
strictly positive functors and indexed inductive linear types.  We
interpret a strictly positive functor as a \emph{dependent polynomial
functor} on the category of sets, also sometimes called an
\emph{indexed container} \cite{gambino_wellfounded_2004,altenkirch_indexed_2015}.
\begin{definition}
  Let $I$ and $O$ be sets. A (dependent) \emph{polynomial} (of sets)
  from $I$ to $O$ consists of a set of shapes $S$, a set of positions
  $P$ and functions $f : P \to I$, $g : P \to S$ and $h : S \to O$. The \emph{extension}
  of a polynomial is a functor $\Set/I \to \Set/O$ defined as the
  composite
  % https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXFNldC9JIl0sWzMsMCwiXFxTZXQvTyJdLFsxLDAsIlxcU2V0L1AiXSxbMiwwLCJcXFNldC9TIl0sWzAsMiwiZl4qIl0sWzIsMywiXFxQaV9nIl0sWzMsMSwiXFxTaWdtYV9oIl1d
  \[\begin{tikzcd}
	    {\Set/I} & {\Set/P} & {\Set/S} & {\Set/O}
	    \arrow["{f^*}", from=1-1, to=1-2]
	    \arrow["{\Pi_g}", from=1-2, to=1-3]
	    \arrow["{\Sigma_h}", from=1-3, to=1-4]
  \end{tikzcd}\]
  Where $f^*$ is the pullback functor along $f$; $\Pi_g$ is the dependent
  product operation and $\Sigma_h$ is the dependent sum operation, which
  are, respectively, the right and left adjoint of their pullback functors
  $g^*$ and $h^*$. A dependent polynomial functor from $I$ to $O$ is a functor $\Set/I
  \to \Set/O$ that is naturally isomorphic to the extension of a
  dependent polynomial from $I$ to $O$.
\end{definition}
With this interpretation of $F$ as a polynomial functor, $\el(F)$ and
$\map(F)$ are just interpreted as the action of the functor on objects
and morphisms, respectively. We interpret the constructors
$\mathsf{K},\Var$, etc. on functors in the obvious way that matches
the definitional behavior of $\el$ and $\map$. The non-trivial part of
the construction is verifying that such such constructions are closed
under being polynomial. The details are tedious but straightforward
extension of prior work on dependent polynomials and indexed
containers and we have verified the construction in Agda.

@@

\begin{center}
\begin{tikzcd}
  \semg{\Delta,\Delta'} \arrow[r, "m_{\Delta,\Delta'}"] &
  \semg{\Delta} \otimes \semg{\Delta'} \arrow[r, "\semg{e_1} \otimes \semg{e_2}"] &
  \semg{A} \otimes \semg{G}
\end{tikzcd}
\end{center}

@@

\begin{center}
\begin{tikzcd}
  \semg{\Delta,\Delta'} \arrow[r , "m_{\Delta,\Delta'}"] &
  \semg{\Delta} \otimes \semg{\Delta'} \arrow[r , "id \otimes \semg{e}"] &
  \semg{\Delta} \otimes \semg{A} \arrow[r , "\phi^{-1}\left( \semg{e'} \right)"]
  &
  \semg{B}
\end{tikzcd}
\end{center}

@@

\begin{center}
\begin{tikzcd}
  \semg{A} \otimes \semg{\Delta} \arrow[r , "\lambda^{-1} \otimes id"] &
  \semg{a : A} \otimes \semg{\Delta} \arrow[r , "m^{-1}_{a : A , \Delta}"] &
  \semg{a : A , \Delta} \arrow[r , "\semg{e}"] &
  \semg{B}
\end{tikzcd}
\end{center}

@@

\begin{center}
\begin{tikzcd}
  \semg{\Delta',\Delta} \arrow[r , "m_{\Delta',\Delta}"] &
  \semg{\Delta'} \otimes \semg{\Delta} \arrow[r , "\semg{e'} \otimes id"] &
  \semg{A} \otimes \semg{\Delta} \arrow[r , "\psi^{-1}\left( \semg{e} \right)"] &
  \semg{B}
\end{tikzcd}
\end{center}

@@

\subsubsection{$\bigamp$-Products}
% Products
\paragraph{$\bigamp$-Introduction}
$\semg{\dlamb x e} : \semg{\Delta} \to \prod_{x : X} {\sem{A}(\gamma , x)}$ is
defined by the universal property of the product
\[
\semg{\dlamb x e} = \left( \sem{e}(\gamma , x) \right)_{(x : X)}
\]

@@

\begin{center}
\begin{tikzcd}
  \semg{\Delta} \arrow[r , "\semg{e}"] &
  \prod_{x : X} {\sem{A}(\gamma , x)} \arrow[r , "\pi_M"] &
  \sem{A}(\gamma , M)
\end{tikzcd}
\end{center}

\paragraph{$\bigamp\beta$}
The $\beta$ law for $\bigamp$ is given by,
\[
  \sem{(\dlamb x e).\pi~M}(\gamma, x) = \sem{e[M/x]}(\gamma, x)
\]
\begin{proof}
\begin{align*}
  \semg{(\dlamb x e).\pi~M}
  & = \pi_{M} \circ \semg{\dlamb x e} \\
  & = \pi_{M} \circ (\sem{e}(\gamma , x))_{(x : X)} \\
  & = \sem{e}(\gamma , M)
\end{align*}
by the universal property of the product.

@@

\paragraph{$\bigamp\eta$}
The $\eta$ law for $\bigamp$ is given by,
\[
  \semg{(\dlamb x {e.\pi~x})} = \semg{e}
\]
\begin{proof}
\begin{align*}
  \semg{(\dlamb x {e.\pi~x})}
  & = (\semg{e.\pi~x})_{x : X} \\
  & = (\pi_x \circ \semg{e})_{x : X} \\
  & = \semg{e}
\end{align*}
\end{proof}
by the universal property of the product.

\subsubsection{$\bigoplus$-Sums}
% Sums
\paragraph{$\bigoplus$-Introduction}
$\semg{\sigma~M~e} : \semg{\Delta} \to \coprod_{x : X} \sem{A}(\gamma, x)$

\begin{center}
\begin{tikzcd}
  \semg{\Delta} \arrow[r , "\semg{e}"] &
  \sem{A}(\gamma, M) \arrow[r , "i_M"] &
  \coprod_{x : X} {\sem{A}(\gamma , x)}
\end{tikzcd}
\end{center}

@@

\begin{center}
% https://tikzcd.yichuanshen.de/#N4Igdg9gJgpgziAXAbVABwnAlgFyxMJZABgBpiBdUkANwEMAbAVxiRAB124YBbAc2CcAIjAY46AcgD6ARlLDR4+exFjJUgEwBfEFtLpMufIRRzK1es1aIOXXgIVrpcx+K0ACThDw94nu-yCKorq2rr6IBjYeAREcjJUtIwsbJwMMABmOAAU-tyBruoyHl4+fpz5DsFqJewATlh8ABY4AJT+3li+cHn2Qari0mF6BtHGcaQaiZYpNpwAMlhgAMqNACoAnos8uO7AAB4ewAAaRxW8-SHOpHTuiO4AgsoDoVrZnHx0PDy3pO77rR0I0ihhiJhIk2myWstkWK3WWy6uwOR1OezSmRyvR4lycsjeHy+P3cfwBHTKcHYYHcNNp5xxDwJ7E+31+-3anAazTa5K65QCuMGmiZLOJpMB4VGRliKDICQs0NS7HSWVy9KqL2ctU63X8cNWfE222Rhz2aKC+WAjPezKJbIBtS5LQ57B1-MqgtekpBYxlyDkAGYoVYlZaAMIiu3i71RaXgshBhUhuYBAQR3SJGBQPjwIigDJ1CA8JBkEA4CBIORJZMgHhSYA2zWyZ4hVotvFhagMOgAI1EAAVQeMbCqcN6C0XK9Ry0gNEnZrX64VroUtKU+T0sFAQF3ewOhzKQKPx4Xi4g52WK4gAKzzmHZLe83Xq4AwLQut2b7e7vsMQe+kwj0xE9J0QANpyvAA2O82GQbFXwkSNWRJdl3Aoetsn2O53GOCUf33AC2GPYEJzPAAWCCkHA6sFx4AA9YAAFpigw5dmzoRAHjbNjOyPPc-wPQDiIiUikFvS8kAomiYR3Pjf3-OMiOAkjTyQaCJMQAB2LQKC0IA
\begin{tikzcd}[column sep =large]
{\semg{\Delta'_1,\Delta,\Delta'_2}} \arrow[r, "{m_{(\Delta'_1,\Delta),\Delta'_2}}"]                                                                                                           & {\semg{\Delta'_1,\Delta} \otimes \semg{\Delta'_2}} \arrow[d, "{m_{\Delta'_1,\Delta}\otimes id}"]                             \\
{\left( \semg{\Delta'_1} \otimes \coprod_{x:X} {\sem{A}(\gamma , x)} \right) \otimes \semg{\Delta'_2}} \arrow[d, "d"]                                                                 & \left( \semg{\Delta'_1} \otimes \semg{\Delta} \right) \otimes \semg{\Delta'_2} \arrow[l, "(id \otimes \semg{e}) \otimes id"] \\
{\coprod_{x : X} {\left( \sem{\Delta'_1}(\gamma , x) \otimes
      \sem{A}(\gamma , x) \right) \otimes \sem{\Delta'_2}(\gamma , x)}} \arrow[r, "\coprod_{x:X}({m^{-1}_{(\Delta'_1,a:A),\Delta'_2}})"] & {\coprod_{x : X} {\sem{\Delta'_1,a : A,\Delta'_2}(\gamma , x)}} \arrow[d, "{[ \sem{e'}(\gamma , x) ]_{(x : X)}}"]     \\
\semg{C}                                                                                                                                                                                      & {\sem{C}(\gamma, x)} \arrow[l]
\end{tikzcd}
\end{center}

@@

\paragraph{$\bigoplus\beta$}
The $\beta$ rule for $\bigoplus$ is given by,
\[
  \semg{\letin {\sigma~x~a} {\sigma~M~a'} {e'}} = \semg{e'[M/x,a'/a]}
\]
\begin{proof}
  \begin{align*}
  \llbracket \mathsf{let}\, {\sigma~x~a} = & {\sigma~M~a'} \, \mathsf{in}\, {e'} \rrbracket \gamma \\
  % \semg{\letin {\sigma~x~a} {\sigma~M~a'} {e'}}
  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ \coprod_{x:X}(m^{-1}) \circ d \circ (id
  \otimes \semg{\sigma~M~a'}) \otimes id \circ m \otimes id \circ m \\
  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ \coprod_{x:X}m^{-1} \circ d \circ (id
  \otimes (i_M \circ \semg{a'})) \otimes id \circ m \otimes id \circ m \\
  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ \coprod_{x:X} m^{-1} \circ d \circ (id \otimes i_M) \otimes id \circ (id \otimes  \lambda) \otimes id \circ m \otimes id \circ m \\
  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ \coprod_{x:X} m^{-1} \circ i_M \circ (id \otimes  \lambda) \otimes id \circ m \otimes id \circ m \\
  & = \sem{e'}(\gamma , M) \circ m^{-1} (id \otimes  \lambda) \otimes id \circ m \otimes id \circ m \\
  & = \sem{e'}(\gamma , M) \tag{coherence}
  \end{align*}

@@

\paragraph{$\bigoplus\eta$}
It suffices to show
\[
  \semg{\letin {\sigma~x~a} {c'} {f[(\sigma~x~a)/c]}} = \semg{f[c'/c]} = \semg{f}
  \]
First, expanding the left hand side, we have.
% \newcommand{\sem}[1]{\left\llbracket{#1}\right\rrbracket}
% \newcommand{\semg}[1]{\sem{#1}\gamma}
% \newcommand{\letin}[3]{\mathsf{let}\, #1 = #2 \, \mathsf{in}\, #3}
\begin{proof}
  \begin{align*}
    \llbracket \mathsf{let}\, {\sigma~x~a} = {c'} & \, \mathsf{in}\, {f[(\sigma~x~a)/c]} \rrbracket \gamma \\
    &= [ \semg{f[(\sigma~x~a)/c]} ]_{(x : X)} \circ \coprod_{x:X}(m^{-1}) \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
    &= [ \semg{f} \circ (id \otimes i_x) \otimes id ]_{(x : X)} \circ \coprod_{x:X}(m^{-1}) \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
    &= \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
  \end{align*}
Since the domain has the universal property of a coproduct (due to distributivity), to prove
this is equal to $\semg{f}$, it is sufficient to prove they are equal
when composed with the injections:

\begin{align*}
  \semg{f} &\circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \circ (id \otimes i_y) \otimes id\\
  &= \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ d  \circ (id \otimes i_y) \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \tag{naturality}\\
  &= \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ i_y \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \tag{naturality}\\
  &= \semg{f} \circ (id \otimes i_y) \otimes id \circ (m^{-1}) \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \\
  &= \semg{f} \circ (m^{-1}) \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \circ (id \otimes i_y) \otimes id \\
  &= \semg{f} \circ (id \otimes i_y) \otimes id \tag{coherence}\\
\end{align*}
\end{proof}

>>>>

Next, a parse in the equalizer $\equalizer{a}{f}{g}$ is defined as
a parse in $\sem{A}$ that is mapped to the same parse by the parse transformers $\sem{f}$ and
$\sem{g}$.
%
The universe $L$ of linear types is interpreted as the set of all
small grammars.
\begin{figure}
  \begin{minipage}[t]{.3\textwidth}
  \begin{footnotesize}
    \begin{flalign*}
    &\sem{c}\gamma\,w = \{ c | w = c\}\\
    &\sem{I}\gamma\,w = \{ () \pipe w = \epsilon \}\\
    &\sem{\uparrow A}\gamma = \sem{A}\gamma\,\epsilon\\
    &\sem{A \lto B}\gamma\,w = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,ww'\\
    &\sem{B \tol A}\gamma\,w = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'w\\
    &\sem{\el(F)}\gamma\,G = \sem{F}\gamma\,G\\
    &\sem{\map(F)}\gamma\,f = \sem{F}\gamma\,f\\
    &\sem{\mu A}\gamma = \mu (\sem{A}\gamma)
  \end{flalign*}
  \end{footnotesize}
  \end{minipage}%
  \begin{minipage}[t]{.7\textwidth}
  \begin{footnotesize}
    \begin{flalign*}
    &\sem{A \otimes B}\gamma\,w = \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in \sem{A}\gamma\,w_1 \wedge b \in \sem{B}\gamma\,w_2 \}\\
    % &\sem{\bigoplus_{x:X} A}\gamma\,w = \{ (x, a) \pipe x \in \sem{X}\gamma \wedge a \in \sem{A}(\gamma,x)w \} \\
    &\sem{\bigoplus_{x:X} A}\gamma\,w = \coprod_{x \in \sem{X}\gamma} \sem{A}(\gamma, x)\,w \\
    &\sem{\bigwith_{x:X} A}\gamma\,w = \prod_{x\in \sem{X}\gamma} \sem{A}(\gamma, x)\,w \\
    &\sem{\equalizer{a}{f}{g}}\gamma\,w = \{ a \in \sem{A}\gamma\,w \pipe \sem{f}\gamma\,w\,a = \sem{g}\gamma\,w\,a \} \\
    &\sem{L}\gamma = \Grammar_0\\
    &\sem{SPF\,X}\gamma = \textrm{DepPolyFunctor}(\sem{X}\gamma\times \Sigma^*,\Sigma^*)
  \end{flalign*}
  \end{footnotesize}
  \end{minipage}
  \caption{Grammar Semantics}
  \label{fig:semantics}
\end{figure}
The most complex part of the semantics is the interpretation of
strictly positive functors and indexed inductive linear types.  We
interpret a strictly positive functor as a \emph{dependent polynomial
functor} on the category of sets, also sometimes called an
\emph{indexed container} \cite{gambino_wellfounded_2004,altenkirch_indexed_2015}.
\begin{definition}
  Let $I$ and $O$ be sets. A (dependent) \emph{polynomial} (of sets)
  from $I$ to $O$ consists of a set of shapes $S$, a set of positions
  $P$ and functions $f : P \to I$, $g : P \to S$ and $h : S \to O$. The \emph{extension}
  of a polynomial is a functor $\Set/I \to \Set/O$ defined as the
  composite
  % https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXFNldC9JIl0sWzMsMCwiXFxTZXQvTyJdLFsxLDAsIlxcU2V0L1AiXSxbMiwwLCJcXFNldC9TIl0sWzAsMiwiZl4qIl0sWzIsMywiXFxQaV9nIl0sWzMsMSwiXFxTaWdtYV9oIl1d
  \[\begin{tikzcd}
	    {\Set/I} & {\Set/P} & {\Set/S} & {\Set/O}
	    \arrow["{f^*}", from=1-1, to=1-2]
	    \arrow["{\Pi_g}", from=1-2, to=1-3]
	    \arrow["{\Sigma_h}", from=1-3, to=1-4]
  \end{tikzcd}\]
  Where $f^*$ is the pullback functor along $f$; $\Pi_g$ is the dependent
  product operation and $\Sigma_h$ is the dependent sum operation, which
  are, respectively, the right and left adjoint of their pullback functors
  $g^*$ and $h^*$. A dependent polynomial functor from $I$ to $O$ is a functor $\Set/I
  \to \Set/O$ that is naturally isomorphic to the extension of a
  dependent polynomial from $I$ to $O$.
\end{definition}
With this interpretation of $F$ as a polynomial functor, $\el(F)$ and
$\map(F)$ are just interpreted as the action of the functor on objects
and morphisms, respectively. We interpret the constructors
$\mathsf{K},\Var$, etc. on functors in the obvious way that matches
the definitional behavior of $\el$ and $\map$. The non-trivial part of
the construction is verifying that such such constructions are closed
under being polynomial. The details are tedious but straightforward
extension of prior work on dependent polynomials and indexed
containers and we have verified the construction in Agda.

@@

\begin{center}
\begin{tikzcd}[column sep = 1in]
  \semg{\Delta,\Delta'} \arrow[r, "m_{\Delta,\Delta'}"] &
  \semg{\Delta} \otimes \semg{\Delta'} \arrow[r, "\semg{e_1} \otimes \semg{e_2}"] &
  \semg{A} \otimes \semg{G}
\end{tikzcd}
\end{center}

@@

\begin{center}
\begin{tikzcd}[column sep=.75in]
  \semg{\Delta,\Delta'} \arrow[r , "m_{\Delta,\Delta'}"] &
  \semg{\Delta} \otimes \semg{\Delta'} \arrow[r , "id \otimes \semg{e}"] &
  \semg{\Delta} \otimes \semg{A} \arrow[r , "\phi^{-1}\left( \semg{e'} \right)"]
  &
  \semg{B}
\end{tikzcd}
\end{center}

@@

\begin{center}
\begin{tikzcd}[column sep=.75in]
  \semg{A} \otimes \semg{\Delta} \arrow[r , "\lambda^{-1} \otimes id"] &
  \semg{a : A} \otimes \semg{\Delta} \arrow[r , "m^{-1}_{a : A , \Delta}"] &
  \semg{a : A , \Delta} \arrow[r , "\semg{e}"] &
  \semg{B}
\end{tikzcd}
\end{center}

@@

\begin{center}
\begin{tikzcd}[column sep=.75in]
  \semg{\Delta',\Delta} \arrow[r , "m_{\Delta',\Delta}"] &
  \semg{\Delta'} \otimes \semg{\Delta} \arrow[r , "\semg{e'} \otimes id"] &
  \semg{A} \otimes \semg{\Delta} \arrow[r , "\psi^{-1}\left( \semg{e} \right)"] &
  \semg{B}
\end{tikzcd}
\end{center}

@@

\subsubsection{$\bigamp$-Products}
% Products
\paragraph{$\bigamp$-Introduction}
$\semg{\dlamb x e} : \semg{\Delta} \to \prod_{x : \semg{X}} {\sem{A}(\gamma , x)}$ is
defined by the universal property of the product
\[
\semg{\dlamb x e} = \left( \sem{e}(\gamma , x) \right)_{(x : \semg{X})}
\]

@@

\begin{center}
\begin{tikzcd}
  \semg{\Delta} \arrow[r , "\semg{e}"] &
  \prod_{x : \semg{X}} {\sem{A}(\gamma , x)} \arrow[r , "\pi_M"] &
  \sem{A}(\gamma , M)
\end{tikzcd}
\end{center}

\paragraph{$\bigamp\beta$}
The $\beta$ law for $\bigamp$ is given by,
\[
  \sem{(\dlamb x e).\pi~M}(\gamma, x) = \sem{e[M/x]}(\gamma, x)
\]
\begin{proof}
\begin{align*}
  \semg{(\dlamb x e).\pi~M}
  & = \pi_{M} \circ \semg{\dlamb x e} \\
  & = \pi_{M} \circ (\sem{e}(\gamma , x))_{(x : \semg{X})} \\
  & = \sem{e}(\gamma , M)
\end{align*}
by the universal property of the product.

@@

\paragraph{$\bigamp\eta$}
The $\eta$ law for $\bigamp$ is given by,
\[
  \semg{(\dlamb x {e.\pi~x})} = \semg{e}
\]
\begin{proof}
\begin{align*}
  \semg{(\dlamb x {e.\pi~x})}
  & = (\semg{e.\pi~x})_{x : \semg{X}} \\
  & = (\pi_x \circ \semg{e})_{x : \semg{X}} \\
  & = \semg{e}
\end{align*}
\end{proof}
by the universal property of the product.

\subsubsection{$\bigoplus$-Sums}
% Sums
\paragraph{$\bigoplus$-Introduction}
$\semg{\sigma~M~e} : \semg{\Delta} \to \coprod_{x : \semg{X}} \sem{A}(\gamma, x)$

\begin{center}
\begin{tikzcd}
  \semg{\Delta} \arrow[r , "\semg{e}"] &
  \sem{A}(\gamma, M) \arrow[r , "i_M"] &
  \coprod_{x : \semg{X}} {\sem{A}(\gamma , x)}
\end{tikzcd}
\end{center}

@@

\begin{center}
% https://tikzcd.yichuanshen.de/#N4Igdg9gJgpgziAXAbVABwnAlgFyxMJZABgBpiBdUkANwEMAbAVxiRAB124YBbAc2CcAIjAY46AcgD6ARlLDR4+exFjJUgEwBfEFtLpMufIRRzK1es1aIOXXgIVrpcx+K0ACThDw94nu-yCKorq2rr6IBjYeAREcjJUtIwsbJwMMABmOAAU-tyBruoyHl4+fpz5DsFqJewATlh8ABY4AJT+3li+cHn2Qari0mF6BtHGcaQaiZYpNpwAMlhgAMqNACoAnos8uO7AAB4ewAAaRxW8-SHOpHTuiO4AgsoDoVrZnHx0PDy3pO77rR0I0ihhiJhIk2myWstkWK3WWy6uwOR1OezSmRyvR4lycsjeHy+P3cfwBHTKcHYYHcNNp5xxDwJ7E+31+-3anAazTa5K65QCuMGmiZLOJpMB4VGRliKDICQs0NS7HSWVy9KqL2ctU63X8cNWfE222Rhz2aKC+WAjPezKJbIBtS5LQ57B1-MqgtekpBYxlyDkAGYoVYlZaAMIiu3i71RaXgshBhUhuYBAQR3SJGBQPjwIigDJ1CA8JBkEA4CBIORJZMgHhSYA2zWyZ4hVotvFhagMOgAI1EAAVQeMbCqcN6C0XK9Ry0gNEnZrX64VroUtKU+T0sFAQF3ewOhzKQKPx4Xi4g52WK4gAKzzmHZLe83Xq4AwLQut2b7e7vsMQe+kwj0xE9J0QANpyvAA2O82GQbFXwkSNWRJdl3Aoetsn2O53GOCUf33AC2GPYEJzPAAWCCkHA6sFx4AA9YAAFpigw5dmzoRAHjbNjOyPPc-wPQDiIiUikFvS8kAomiYR3Pjf3-OMiOAkjTyQaCJMQAB2LQKC0IA
\begin{tikzcd}[column sep =large]
{\semg{\Delta'_1,\Delta,\Delta'_2}} \arrow[r, "{m_{(\Delta'_1,\Delta),\Delta'_2}}"]                                                                                                           & {\semg{\Delta'_1,\Delta} \otimes \semg{\Delta'_2}} \arrow[d, "{m_{\Delta'_1,\Delta}\otimes id}"]                             \\
{\left( \semg{\Delta'_1} \otimes \coprod_{x:\semg{X}} {\sem{A}(\gamma , x)} \right) \otimes \semg{\Delta'_2}} \arrow[d, "d"]                                                                 & \left( \semg{\Delta'_1} \otimes \semg{\Delta} \right) \otimes \semg{\Delta'_2} \arrow[l, "(id \otimes \semg{e}) \otimes id"] \\
{\coprod_{x : \semg{X}} {\left( \sem{\Delta'_1}(\gamma , x) \otimes
      \sem{A}(\gamma , x) \right) \otimes \sem{\Delta'_2}(\gamma , x)}} \arrow[r, outer sep=.15cm, "\coprod_{x:\semg{X}}({m^{-1}_{(\Delta'_1,a:A),\Delta'_2}})"] & {\coprod_{x : \semg{X}} {\sem{\Delta'_1,a : A,\Delta'_2}(\gamma , x)}} \arrow[d, "{[ \sem{e'}(\gamma , x) ]_{(x : \semg{X})}}"]     \\
\semg{C}                                                                                                                                                                                      & {\sem{C}(\gamma, x)} \arrow[l]
\end{tikzcd}
\end{center}

@@

\paragraph{$\bigoplus\beta$}
The $\beta$ rule for $\bigoplus$ is given by,
\[
  \semg{\letin {\sigma~x~a} {\sigma~M~a'} {e'}} = \semg{e'[M/x,a'/a]}
\]
\begin{proof}
  \begin{align*}
  \llbracket \mathsf{let}\, {\sigma~x~a} = & {\sigma~M~a'} \, \mathsf{in}\, {e'} \rrbracket \gamma \\
  % \semg{\letin {\sigma~x~a} {\sigma~M~a'} {e'}}
  & = {[ \sem{e'}(\gamma , x) ]_{(x : \semg{X})}} \circ \coprod_{x:\semg{X}}(m^{-1}) \circ d \circ (id
  \otimes \semg{\sigma~M~a'}) \otimes id \circ m \otimes id \circ m \\
  & = {[ \sem{e'}(\gamma , x) ]_{(x : \semg{X})}} \circ \coprod_{x:\semg{X}}m^{-1} \circ d \circ (id
  \otimes (i_M \circ \semg{a'})) \otimes id \circ m \otimes id \circ m \\
  & = {[ \sem{e'}(\gamma , x) ]_{(x : \semg{X})}} \circ \coprod_{x:\semg{X}} m^{-1} \circ d \circ (id \otimes i_M) \otimes id \circ (id \otimes  \lambda) \otimes id \circ m \otimes id \circ m \\
  & = {[ \sem{e'}(\gamma , x) ]_{(x : \semg{X})}} \circ \coprod_{x:\semg{X}} m^{-1} \circ i_M \circ (id \otimes  \lambda) \otimes id \circ m \otimes id \circ m \\
  & = \sem{e'}(\gamma , M) \circ m^{-1} (id \otimes  \lambda) \otimes id \circ m \otimes id \circ m \\
  & = \sem{e'}(\gamma , M) \tag{coherence}
  \end{align*}

@@

\paragraph{$\bigoplus\eta$}
It suffices to show
\[
  \semg{\letin {\sigma~x~a} {c'} {f[(\sigma~x~a)/c]}} = \semg{f[c'/c]} = \semg{f}
  \]
First, expanding the left hand side, we have.
% \newcommand{\sem}[1]{\left\llbracket{#1}\right\rrbracket}
% \newcommand{\semg}[1]{\sem{#1}\gamma}
% \newcommand{\letin}[3]{\mathsf{let}\, #1 = #2 \, \mathsf{in}\, #3}
\begin{proof}
  \begin{align*}
    \llbracket \mathsf{let}\, {\sigma~x~a} = {c'} & \, \mathsf{in}\, {f[(\sigma~x~a)/c]} \rrbracket \gamma \\
    &= [ \semg{f[(\sigma~x~a)/c]} ]_{(x : \semg{X})} \circ \coprod_{x:\semg{X}}(m^{-1}) \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
    &= [ \semg{f} \circ (id \otimes i_x) \otimes id ]_{(x : \semg{X})} \circ \coprod_{x:\semg{X}}(m^{-1}) \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
    &= \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : \semg{X})} \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
  \end{align*}
Since the domain has the universal property of a coproduct (due to distributivity), to prove
this is equal to $\semg{f}$, it is sufficient to prove they are equal
when composed with the injections:

\begin{align*}
  \semg{f} &\circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : \semg{X})} \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \circ (id \otimes i_y) \otimes id\\
  &= \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : \semg{X})} \circ d  \circ (id \otimes i_y) \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \tag{naturality}\\
  &= \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : \semg{X})} \circ i_y \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \tag{naturality}\\
  &= \semg{f} \circ (id \otimes i_y) \otimes id \circ (m^{-1}) \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \\
  &= \semg{f} \circ (m^{-1}) \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \circ (id \otimes i_y) \otimes id \\
  &= \semg{f} \circ (id \otimes i_y) \otimes id \tag{coherence}\\
\end{align*}
\end{proof}
